---
title: "GENIE3 Guide"
author: "Fang Shu"
date: "August 2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This guide guides the user in using the R package GENIE3
```{r}
data<-read.csv("PreprocessedExpressionData_transposed.csv", header = TRUE)
data1<-data.frame(data[,-1],row.names = data[,1])
dim(data1)
```
Since the package GENIE3 is not in the R Cran Repository, use the following code to download
```{r}
#if(!requireNamespace("BiocManager",quietly = TRUE))
#install.packages("BiocManager")  
#BiocManager::install("GENIE3")
```

```{r}
library(GENIE3)
data2<-t(data1)
dim(data2)
```
Run GENIE3 with its default parameters.
GENIE3 infers a gene regulatory network in the form of a weighted adjacency matrix from expression data,using ensembles of regression trees.
The input is an expression matrix, where each row is a gene and every column is a sample.

```{r}
weightMat<-GENIE3(data2)#The weighted adjacency matrix
weightMat[1:10,1:6]
dim(weightMat)
```
This command computes the weighted adjacency matrix of the gene network with the GENIE3 algorithm (Random Forests).In the weight
matrix, element (i,j) (row i, column j) gives the weight of the link from regulatory gene i to target gene j, with high
scores corresponding to more likely regulatory links.
The algorithm outputs a matrix containing the weights of the putative regulatory links, with higher weights corresponding to
more likely regulatory links. weightMat[i,j] is the weight of the link directed from the i-th gene to j-th gene.


Restrict the candidate regulators to a subset of genes
By default, all the genes are used as candidate regulators. The list of candidate regulators can however be restricted to a 
subset of genes. This can be useful when you know which genes are transcription factors.

```{r}
regulators <- c("MKL2","CD109","MAST2")
regulators
weightMat <- GENIE3(data2,regulators = regulators)
weightMat[1:3,1:6]
```
Here only Gene MKL2, Gene CD109,and Gene MAST2 were used as candidate regulators. In the resulting weightMat, the links that are directed from genes that are not candidate regulators have a weight equal to 0.

Change the tree-based method and its settings

GENIE3 is based on regression trees. These trees can be learned using either the Random Forest method or the Extra-Trees method.
The tree-based method can be specified using the tree.method parameter(tree.method="RF") for Random Forests,or tree.method="ET" for Extra-Trees.
Each tree-based method has 2 parameters: K and ntrees. K is the number of candidate regulators that are randomly selected at 
each tree node for the best split determination. P is the number of candidate regulators. K must be between 1 and p.
The parameter ntrees specifies the number of trees that are grown per ensemble. It can be set to any strictly positive integer.

Use Extra-Trees (ET) method

7 randomly chosen candidate regulators at each node of a tree
50 trees per ensemble
```{r}
weightMat1 <- GENIE3(data2, treeMethod = "ET", K=7,nTrees = 50)
weightMat1[1:10,1:6]
```
Increasing the value of K
```{r}
weightMat2 <-GENIE3(data2,treeMethod = "ET", K=8, nTrees=50)
weightMat2[1:12,1:6]
```
Increasing the value of trees
```{r}
weightMat3 <-GENIE3(data2,treeMethod = "ET", K=7, nTrees = 60)
weightMat3[1:10,1:6]
```

Comparing the effects on the weights by changing K and nTrees
```{r}
weightMat1[weightMat1>weightMat2]<- 1
table(weightMat1==1)
table(weightMat1==0)
weightMat1 <- GENIE3(data2, treeMethod = "ET", K=7,nTrees = 50)
weightMat1[weightMat1>weightMat3]<-1
table(weightMat1==1)
table(weightMat1==0)
```


Use Random-Forest (RF) method
```{r}
weightMat4 <-GENIE3(data2, treeMethod = "RF",K=7, nTrees = 50)
weightMat4[1:10,1:6]
```
Increasing the value of K
```{r}
weightMat5 <-GENIE3(data2, treeMethod = "RF",K=8, nTrees = 50)
weightMat5[1:10,1:6]
```
Increasing the value of trees
```{r}
weightMat6 <-GENIE3(data2, treeMethod = "RF",K=7, nTrees = 60)
weightMat6[1:10,1:6]
```

Comparing the effects on the weights by changing K and nTrees
```{r}
weightMat4[weightMat4>weightMat5]<- 1
table(weightMat4==1)
table(weightMat4==0)
weightMat4 <-GENIE3(data2, treeMethod = "RF",K=7, nTrees = 50)
weightMat4[weightMat4>weightMat6]<-1
table(weightMat4==1)
table(weightMat4==0)
```

INPUT

K is the number of candidate regulators randomly selected at each tree node for the determination of the best split.
Must be either "sqrt" for the square root of the total number of candidate regulators (default), "all" for the total number of candidate regulators, or a stritly positive integer.

nTrees is the number of trees in an ensemble for each target gene, default is 1000.

Get the list of the regulatory links
Get all the regulatory links
Can obtain the list of all the regulatory links (from most likely to least likely):

```{r}
weightMat<-GENIE3(data2)
linkList <- getLinkList(weightMat)#getLinkList converts the weight matrix returned by GENIE3 to a sorted list of regulatory links
dim(linkList)
head(linkList)
```

The resulting linkList matrix contains the ranking of links. Each row corresponds to a regulatory link. The first column shows
the regulator, the second column shows the target gene, and the last column indicates the weight of the link.

Get only the top-ranked links
Usually only interested in extracting the most likely regulatory links.The optional parameter report.max sets the number of 
top-ranked links to report:

```{r}
linkList <- getLinkList(weightMat,reportMax = 5)
linkList
```

Get only the links with a weight higher than some threshold.
A threshold can be set on the weight of the links:
```{r}
linkList <- getLinkList(weightMat,threshold=0.1)
linkList
```

Important note on the interpretation of the weights

The weights of the links returned by GENIE3() do not have any statistical meaning and only provide a way to rank the regulatory
links. There is therefore no standard threshold value,and caution must be taken when choosing one.