---
title: "LEAPS.Rmd"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library("LEAP")
```
Maximum Absolute Correlation (MAC) Counter Function
Can use Mac_counter() function to calculate the Max Absolute Correlation (MAC) matrix for the data.
The output is a matrix where Row gene index and Column gene index correspond to the indices for the gene pair (i,j).
Correlation is the maximum absolute correlation achieved for the pair, and Lag is the lag at which the MAC occured.

Parameters
Setting MAC_cutoff=0.15 restricts the output to only those pairs with an MAC of 0.15 or greater.It is the lowest MAC value
to be shown in the results.
max_lag_prop is the largest proportion of the experiments the user wants the lag to be. The default is 1/3
For this dataset, as there are 200 experiments, the largest lag will be 66
Setting lag_matrix=T creates the lag_example.csv which stores the resulting matrix of lag values
```{r}
data_used <- read.csv("GSE81252.leap.final.new.csv", header = TRUE)
data_used_new <-data.frame(data_used[,-1],row.names = data_used[,1])
dim(data_used_new)
MAC_results = MAC_counter(data = data_used_new, max_lag_prop = 1/3, MAC_cutoff = 0.15,file_name = "example",lag_matrix = T)
MAC_results[99:119,]
MAC_results[1:4,1:4]
data2_new <- read.csv("MAC_example.csv")
dim(data2_new)
data2_new[1:4,1:4]
```

Note the diagonal is set to NA. The lag_example file contains the resulting lag matrix from applying MAC_counter() to the dataset.
```{r}
data3_new <- read.csv("lag_example.csv")
dim(data3_new)
```
Each of the values in the lag matrix correspond to the size of the lag used on the gene listed in the column.
In this example, 23 corresponds to gene 7's expression at its first pseudo-tome point and staggering the expression og gene 6 by 11 pseudo-time points
(hence starting at 24)
```{r}
data3_new[1:10,1:10]
```

Permutation Analysis Function
Parameters
MACs_observ is the resulting matrix of MACs from running MAC_counter on the dataset
The variable num_perms determines the number of permutations to use, default is 100.
FDR_cutoffs determines the number of cutoffs to split the domain [0,1] for the correlation, default is 101.
To determine a cutoff for significant MAC values, can use the MAC_perm function
```{r}
MAC_perm(data = data_used_new, MACs_observ = MAC_example, num_perms = 30, max_lag_prop = 1/3, FDR_cutoffs = 101, perm_file_name = "example")
```
The returned dataset has 4 columns. cors is the correlation cutoffs and MACs_observed are the number of observed correlations at that cutoff.
MACs_ave_perm are the average number observed in the permuted datasets at that cutoff,
and fdr is the false discovery rate (FDR) at that cutoff.
```{r}
data4_new<- read.csv("perm_example.csv")
dim(data4_new)
data4_new[74:101,]
```

Can also view the distribution of lags at various correlation cutoffs.
First, must convert the numerical values in lag matrix into categorical variables based on cutoffs the user will designate.
Pull-out non-NA values 
```{r}
cors=c()
lag=c()
for (i in (1:200)){
  cors=c(cors,na.omit(data2_new[,i]))
  lag=c(lag,na.omit(data3_new[,i]))
}
#Then convert the lag values into categorical cutoffs. 
lag_bin=c()
for(i in (1:79600)){
  if(is.na(lag[i])){
    lag_bin[i]=="NA"
  }
  else if(lag[i]==0){
    lag_bin[i]="0"
  }
  else if(lag[i]>0 & lag[i]<=10){
    lag_bin[i]="1-10"
  }
  else  if(lag[i]>10 & lag[i]<=20){
    lag_bin[i]="11-20"
  }
  else if(lag[i]>20 & lag[i]<=30){
    lag_bin[i]="21-30"
  }
  else if(lag[i]>30 & lag[i]<=40){
    lag_bin[i]="31-40"
  }
  else{
    lag_bin[i]=">40"
  }
}
#Then combine the correlation values and lag bins to create a dataframe ggplot2 can use.
data = as.data.frame(cbind(cors,lag,lag_bin))
dim(data)
data$cors = as.numeric(as.character(data$cors))
data$lag_bin = factor(lag_bin, levels =c(">40", "31-40", "21-30", "11-20", "1-10", "0"))
library(ggplot2)
ggplot(data = data, aes(data$cors)) + geom_histogram(binwidth = 0.05, aes(fill = lag_bin)) + scale_y_sqrt() +xlab("Correlation") +ylab("Square Root Counts")+guides(fill=guide_legend(title="Lag")) + ggtitle("Histogram of Correlations")
```