---
title: "PPCOR Documentation"
author: "Fang Shu"
date: "August 2020"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
data<-read.csv("PreprocessedExpressionData_transposed.csv", header = TRUE)
data1<-data.frame(data[,-1],row.names = data[,1])
library(ppcor)
```

Partial correlation is a measure of the strength and direction of a linear relationship between 2 continuous variables,
while controlling for the effect of one or more other continuous variables.
The pcor function will generate a full matrix of pairwise partial correlations controlling for all variables other than the two variablesin the bivariate correlation, along with tests of statistical significance, for all pairs of variables.

```{r}
pcor(x=data1,method = "pearson")$estimate[1:10,1:6]
pcor(x=data1,method = "spearman")$estimate[1:10,1:6]
#pcor(x=data1,method = "kendall")$estimate
pcor(x=data1,method = "pearson")$p.value[1:10,1:6]
pcor(x=data1,method = "pearson")$statistic[1:10,1:6]
pcor(x=data1,method = "pearson")$n
pcor(x=data1,method = "pearson")$gp
pcor(x=data1,method = "pearson")$method
```
This output has six values.

1) estimate, which is a matrix of the partial correlation coefficient between two variables

2) p.value, which is a matrix of the level of statistical significance/p-vallue of the test

3) statistic, which is a matrix of the value of the test statistic

4) n, which is the total number of samples

5) gp, which is the number of given or controlled variables

6) method, which is the used correlation method (Pearson/default, Kendall, Spearman)

The greater the absolute value of the partial correlation coefficient between a pair of variables, the more linearly correlated
the variables are. And if we do an hypothesis test, the null hypothesis would be that no correlation exists between a pair of variables, while the alternative hypothesis would be that a correlation exists between a pair of variables.
We can also see that for those pairs of variables that have a large partial correlation coefficient, they have a small p-value corresponding, which means we have sufficient evidence to reject the null hypothesis and conclude a correlation exists.

```{r}
#colSums(data1)
#sum(colSums(data1)==0)
```


The partial correlation between "MKL2" and "CD109" given "MAST2".

```{r}
pcor.test(x=data1$MKL2,y=data1$CD109,z=data1$MAST2)
pcor.test(x=data1$MKL2,y=data1$CD109,z=data1$MAST2,method = "spearman")
#pcor.test(x=data1$PLEKHN1,y=data1$CD109,z=data1$MAST2,method = "kendall")
```
x and y are called the focal variables, and z is the control variable. There can be more than one control variables

Can also find the partial correlation between two genes given a third gene by:

```{r}
reg1<-lm(data1$MKL2~data1$MAST2)
resid1<-resid(reg1)
reg2<-lm(data1$CD109~data1$MAST2)
resid2<-resid(reg2)
cor(resid1,resid2)#Generate the same result as pcor.test
cor.test(resid1,resid2)#Gives a confidence interval on the correlation coefficient
```

The semi-partial correlation between "MKL2" and "CD109" given "MAST2".

```{r}
spcor.test(x=data1$MKL2,y=data1$CD109,z=data1$MAST2)
spcor.test(x=data1$MKL2,y=data1$CD109,z=data1$MAST2,method = "spearman")
#spcor.test(x=data1$MKL2,y=data1$CD109,z=data1$MAST2,method = "kendall")
```

We can also be interested in computing the semi-partial correlation between two genes given two other genes.
The semi-partial correlation between MKL2 and CD109 given MAST2 and KAT5 is calculated with

```{r}
spcor.test(x=data1$MKL2,y=data$CD109,z=data1[,c("MAST2","KAT5")])
spcor.test(x=data1$MKL2,y=data$CD109,z=data1[,c("MAST2","KAT5")],method = "spearman")
#spcor.test(x=data1$MKL2,y=data$CD109,z=data1[,c("MAST2","KAT5")],method="kendall")
```

Scatterplots show how much one variable is affected by another, while the relationship between two variables is called their correlation
Generating a pairwise comparison scatterplot

```{r}
data3<- data[,c("MKL2","CD109","MAST2")]
pairs(data3,panel = panel.smooth)
data4<- data[,c("UBE2Z","RGPD4","ELF2")]
pairs(data4,panel = panel.smooth)
pairs(data[,c(5,6:15)])
library(corrplot)
cor_matrix <- cor(data[,c(5,6:15)])
corrplot.mixed(cor_matrix, lower = "circle", upper = "number", tl.pos = "lt", diag = "u")
```

Semi-partial correlation is the correlation of two variables with variation from a third variable removed only from the second variable.
The function spcor can calculate the pairwise semi-partial correlations for each pair of variables given others

```{r}
spcor(x=data1,method = "pearson")$estimate[1:10,1:6]
spcor(x=data1,method = "spearman")$estimate[1:10,1:6]
#spcor(x=data1,method = "kendall")$estimate
```