---
title: "SCODE.Rmd"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Optimize linear Ordinary Differential Equation and infer regulatory network from time course data
```{r}
library(MASS)


maxB <- 2.0
minB <- -10.0
```

#system(paste("mkdir", dir, sep=" "))
#fdata<-read.csv("GSE81252.scode.expression.csv",header=TRUE)
#fdata<-data.frame(fdata[,-1],row.names = fdata[,1])
INPUT 1: Gene expression dataset, each row corresponds to a gene and each column corresponds to a cell
200 transcription factors, 100 cells
```{r}
X <- as.matrix(read.table("GSE81252.scode.expression.txt", header = TRUE, sep = "\t"))[1:200,1:100]
dim(X)

W <- matrix(rep(0,200*2),nrow=200,ncol = 2)
dim(W)
Z <- matrix(rep(0,2*100),nrow=2,ncol=100)
dim(Z)
WZ <- matrix(nrow=200, ncol=100)
dim(WZ)
```

read pseudo-time and normalize pseudo-time
INPUT 2:Pseudo-time dataset,first column is the index of a cell/experimental time point,second column is time parameter (normalized from 0.0 to 1.0)
```{r}
pseudotime <- read.table("GSE81252.scode.time.txt", header = FALSE, sep = "\t")[1:100,2]
dim(pseudotime)
max(pseudotime)
min(pseudotime)
#pseudotime <- (as.matrix(pseudotime))/max(as.matrix(pseudotime))
pseudotime<-pseudotime/max(pseudotime)

new_B <- rep(0, 2)
old_B <- rep(0, 2)

#Initialization
RSS <- Inf
for(i in 1:2){#Initialize a diagonal matrix randomly
  new_B[i] <- runif(1, min=minB, max=maxB)
  old_B[i] <- new_B[i]
}

#function to sample Z
sample_Z <- function(){#Z is generated from dz=Bzdt
  for(i in 1:2){
    for(j in 1:100){
      Z[i,j] <<- exp(new_B[i]*pseudotime[j]) + runif(1, min=-0.001, max=0.001)#includes noise
    }
  }
}

#optimize W and B iteratively
for(ite in 1:100){
  #sampling B
  target <- floor(runif(1, min=1, max=3))
  new_B[target] <- runif(1, min=minB, max=maxB)
  
  #for last calculation
  if(ite == 100){
    for(i in 1:2){
      new_B[i] <- old_B[i]
    }
  }
  
  #sample Z from new B
  sample_Z()
  
  #regression
  for(i in 1:200){#W is solution of linear regression
    X.lm <- lm(X[i,] ~ t(Z)-1)
    for(j in 1:2){
      W[i,j] <- X.lm$coefficients[j]
    }
    WZ[i,] <- W[i,] %*% Z
  }
  
  #RSS
  tmp_RSS <- sum((X-WZ)**2)
  if(tmp_RSS < RSS){#If RSS(B^k,W^k)<RSS(B_hat,W_hat)
    RSS <- tmp_RSS
    old_B[target] <- new_B[target]
  }
  else{#end if B^(k+1)
    new_B[target] <- old_B[target]
  }
}
```
```{r}
dir=getwd()
#output RSS
#The residual sum of squares of linear regression
write.table(RSS, paste(dir,"/RSS.txt",sep=""), row.names=F, col.names=F, sep="\t")

#output W
#A GxD matrix corresponding to W of linear regression
write.table(W, paste(dir,"/W.txt",sep=""), row.names=F, col.names=F, sep="\t")

#infer A
B <- matrix(rep(0,2*2), nrow=2, ncol=2)
for(i in 1:2){
  B[i,i] <- new_B[i]
}
invW <- ginv(W)
A <- W %*% B %*% invW

#output A and B
#This is a 200x200 matrix, corresponding to the inferred regulatory network. Aij represents the regulatory relationship from TF j to TF i.
write.table(A, paste(dir,"/A.txt",sep=""), row.names=F, col.names=F, sep="\t")
#A DxD diagonal matrix which corresponds to the optimized paramters of ODE of z, whereas D is the number of z.
write.table(B, paste(dir,"/B.txt",sep=""), row.names=F, col.names=F, sep="\t")
```

Downstream Analysis
Calculation of RSS (RSS.R)
To choose appropriate size of z, recommend to calculate RSS of independent test data
Input 1: GxC matrix of expression data
Input 2: Pseudo-time data
Output: RSS for this data
G: The number of transcription factors, D: The number of z, C: The number of cells

```{r}
X <- as.matrix(read.table("GSE81252.scode.expression.txt", header = TRUE, sep = "\t"))[1:200,1:100]
dir<-getwd()
W <- as.matrix(read.table(paste(dir,"/W.txt",sep="")))
Z <- matrix(rep(0,2*100), nrow=2, ncol=100)
WZ <- matrix(nrow=200, ncol=100)
B <- read.table(paste(dir,"/B.txt",sep=""))

pseudotime <- read.table("GSE81252.scode.time.txt", header = FALSE, sep = "\t")[1:100,2]
pseudotime <- pseudotime/max(pseudotime)

sample_Z <- function(){
  for(i in 1:2){
    for(j in 1:100){
      Z[i,j] <<- exp(B[i,i]*pseudotime[j]) + runif(1, min=-0.001, max=0.001)
    }
  }
}

sample_Z()
for(i in 1:200){
  WZ[i,] <- W[i,] %*% Z
}

RSS_new <- sum((X-WZ)**2)
RSS_new
write.table(RSS_new, paste(dir,"/RSS_new.txt",sep=""), row.names=F, col.names=F, sep="\t")
```


Reconstruction of expressio ndynamics 
Calculate the dynamics from optimzied linear ODE
Input 1: Initial value of x, each row corresponding to a gene. Column 1 is index of a gene, column2 is initial value.
Input 2: A.txt
Output: (G+1)x101 matrix of reconstructed expression dynamicsdata, where G is the number of transcription factors.
The first column corresponds to time paramter (from 0.0 to 1.0 with 0.01 interval). Each row corresponds to each gene
and each column corresponds to each time point.
```{r}
init <- as.matrix(read.table("intial_try.txt", sep="\t"))[1:200,2]
A <- as.matrix(read.table("A.txt", sep="\t"))
tmp <- eigen(A)
U <- tmp$vectors
invU <- solve(U)
l <- tmp$values
mean <- matrix(nrow=200, ncol=201)
tmpt <- (0:200)*0.01
for(i in 1:length(tmpt)){
  t <- tmpt[i]
  eAt <- Re(U %*% diag(exp(l*t)) %*% invU)
  mean[,i] <- eAt %*% init
}
dir=getwd()
write.table(rbind(tmpt, mean), paste(dir,"/reconstructed_dynamics.txt",sep=""), col.names=F, row.names=F,sep="\t")
```